---
title: "A Gentle Introduction to `plot2()`: Enhancing Your `ggplot2` Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{A Gentle Introduction to `plot2()`: Enhancing Your `ggplot2` Workflow}
  %\VignetteEngine{knitr::rmarkdown}
toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 7
) 
# load these, so that function names will get tidyverse links on the web page
library(ggplot2, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
# now load our own package
library(plot2)
exported <- getNamespaceExports("plot2")
reexported <- sort(exported[!exported %in% ls(envir = asNamespace("plot2"))])
```

# Introduction

If you've been using `ggplot2` for a while, you're likely familiar with its strengths and versatility in creating a wide array of visualisations. However, as powerful as `ggplot2` is, it often requires you to define every single detail yourself. Whether it's specifying aesthetics, selecting geoms, or adjusting scales, you're in control of every element. And that's one of the reasons we love `ggplot2`; it offers immense flexibility and precision.

But what if you could streamline this process for many common tasks? Enter `plot2()`, a friendly companion to `ggplot2` designed to reduce the repetitive aspects of plotting without sacrificing the customizability that `ggplot2` is known for. Think of `plot2()` as your plotting assistant, doing a lot of the heavy lifting automatically so you can focus on the fun parts — like exploring your data and finding insights.

In this vignette, we'll take a deep dive into the `plot2()` function and its companion `add_*()` functions. We'll walk through everything from the basics to some of the more advanced features, helping you unlock the full potential of this powerful tool. Whether you're plotting simple bar charts or complex Sankey diagrams, `plot2()` has got you covered.

## The Plot2 Philosophy: Less Typing, More Plotting

Before we dive into the code, let's talk briefly about the philosophy behind `plot2()`. At its core, `plot2()` is designed to make plotting in R more intuitive and less cumbersome. If you've ever been frustrated by having to write out `ggplot()` and `aes()` over and over again, `plot2()` is the answer.

The concept is simple, and the exact opposite of `ggplot2`: give `plot2()` your data, and it will figure out the rest, while enabling many popular plotting options without ever needing to leave this single function. The goal is to get you from data to visualisation with as little friction as possible. And since `plot2()` just returns a `ggplot` object, you can extend it in any way you would with the outcome of `ggplot()`.

This philosophy shines particularly in how `plot2()` simplifies functionalities like faceting, theming, and applying inline transformations. Rather than managing multiple functions and layers, `plot2()` allows you to achieve the same results more directly.

# Getting Started with `plot2()`

Let's start with the basics. The `plot2()` function is a wrapper around `ggplot2` that simplifies many of the tasks you usually have to handle manually. To begin with, you don't even need to specify what kind of plot you want — `plot2()` will make an educated guess based on your data.

## Basic Usage: Let `plot2()` Do the Work

To get started, you can pass your data directly into `plot2()` without specifying any additional arguments. For example:

```{r, echo = FALSE}
library(plot2)
df <- data.frame(group = c("A", "B", "C", "D"),
                 values = c(105, 120, 114, 136))
```

```{r, eval = FALSE}
# Load the package
library(plot2)

df <- data.frame(group = c("A", "B", "C", "D"),
                 values = c(105, 120, 114, 136))
df |>
  plot2()
```

```{r, echo = FALSE}
p1 <- df |> 
  plot2(title = "df |>\n  plot2()") +
  theme(plot.title = element_text(hjust = 0, family = "Fira Code", size = 10))
p2 <- ggplot(df) +
  geom_col(aes(group, values)) +
  labs(title = "ggplot(df) +\n  geom_col(aes(group, values))") +
  theme(plot.title = element_text(hjust = 0, family = "Fira Code", size = 10))
p1 + p2
```

In this simple example, `plot2()` automatically generates a column plot because it recognises that the x-axis is categorical and the y-axis is numeric. It even adds data labels by default because the x-axis is discrete. Also, at the top of the y-axis there is a bit more space to better be able to read the plot, akin to how Microsoft Excel plots at default. This is `plot2()` at its most basic — no need to specify a plot type or worry about details like labels and scales.

You'll also notice the clean, uncluttered appearance of the plot, especially when compared to a default `ggplot2` plot, thanks to `theme_minimal2()`, which is applied by default in `plot2()`. Unlike `ggplot2`’s default `theme_grey()` that features a grey background, `theme_minimal2()` provides an even more spacious, white background that reduces visual clutter and is optimised for printing directly to production formats like PDFs. This makes `plot2()` plots ideal for reports, publications, and presentations where a clean, professional look is essential.

## Customising the Plot Type

But what if you want something different? Maybe a scatter plot or a line chart? No problem. You can easily specify the `type` argument to get exactly what you want.

```{r}
df |> 
  plot2(type = "l")
```

Here, we've explicitly set `type = "l"` to create a line plot. `plot2()` supports a wide range of plot types, and you can use either the full name (`type = "geom_line"`) or an abbreviation (`type = "line"`, `type = "l"`).

This flexibility can save you from remembering multiple `ggplot2` functions like `geom_point()`, `geom_line()`, and `geom_col()`. With `plot2()`, these are all accessible through a simple `type` argument, making your workflow more intuitive.

## A Closer Look at the Axes

One of the key strengths of `plot2()` is how it handles axes. The `x` and `y` arguments are straightforward, but they come with a lot of flexibility. Let’s start with the basics and then build up to more advanced configurations.

### Setting Up Basic Axes

To create a simple scatter plot, you might pass a single variable to the `x` and `y` arguments:

```{r, eval = FALSE}
mtcars |>
  plot2(mpg, hp)
```

```{r, echo = FALSE}
p1 <- mtcars |> 
  plot2(mpg, hp, title = "mtcars |>\n  plot2(mpg, hp)") +
  theme(plot.title = element_text(hjust = 0, family = "Fira Code", size = 10))
p2 <- ggplot(mtcars) +
  geom_point(aes(mpg, hp)) +
  labs(title = "ggplot(mtcars) +\n  geom_point(aes(mpg, hp))") +
  theme(plot.title = element_text(hjust = 0, family = "Fira Code", size = 10))
p1 + p2
```

This creates a scatter plot of miles per gallon (`mpg`) against horsepower (`hp`). This is as basic as it gets — two variables, one for each axis. Also, `plot2()` always tries to start at zero for both `x` and `y` (which can be set with `x.expand` and `y.expand`). With `ggplot()`, the default plot is 'zoomed in' to the data region.

### Working with Multiple Variables

Now, what if you want to compare multiple variables on the same plot? `plot2()` makes this easy by allowing you to pass a vector of variables:

```{r}
mtcars |> 
  plot2(x = mpg, y = c(hp, disp))
```

In this example, `plot2()` plots both `hp` and `disp` against `mpg` on the same graph, using different colours or other visual distinctions to separate them. This functionality replaces the need for pre-processing steps such as `tidyr::pivot_longer()` for the selected variables, simplifying the code significantly.

## Delving into Categories

Categories in `plot2()` replace the `colour` and `fill` aesthetics from `ggplot2`. The `category` argument is where you define how your data should be grouped. Let’s start with a simple example and then explore how to take it further.

### Basic Grouping

A basic use of the `category` argument might look something like this:

```{r}
mtcars |> 
  plot2(mpg, hp, category = cyl)
```

Here, the `cyl` variable is used to group the data by the number of cylinders, and `plot2()` automatically assigns different colours to each group. This is particularly useful for comparing subsets of your data within the same plot.

In `ggplot2`, this would typically involve specifying `aes(colour = ...)` or `aes(fill = ...)` within a `geom` function. `plot2()` simplifies this by handling it directly through the `category` argument, reducing the need for manually setting aesthetics.

### Customising Categories

But `plot2()` doesn’t stop there. You can fully customise how categories are displayed. For instance, you might want to control the colours used for each category:

```{r}
mtcars |> 
  plot2(mpg, hp, category = cyl,
        colour = c("4" = "red", "6" = "blue", "8" = "green"))
```

Here, we’ve specified the exact colours to use for each category, giving you full control over the appearance of your plot.

This approach mirrors what you would do with `scale_colour_manual()` or `scale_fill_manual()` in `ggplot2`, but in a more streamlined and integrated manner.

## Exploring Facets

Faceting in `plot2()` allows you to split your plot into multiple panels, one for each level of a categorical variable. It’s an excellent way to compare data across different groups.

### Basic Faceting

A simple example of faceting might look like this:

```{r}
mtcars |> 
  plot2(mpg, hp, facet = gear)
```

This command splits the plot by the number of gears, giving you a separate panel for each group. This is a quick way to see how relationships vary across different subsets of your data.

In `ggplot2`, achieving this would typically require `facet_wrap(~ gear)` or `facet_grid(gear ~ .)`. With `plot2()`, you achieve the same effect with a simple `facet` argument, making it more intuitive and easier to remember.

### Advanced Faceting Options

`plot2()` offers additional control over how facets are displayed. You can specify the number of rows in the facet grid, control whether scales are fixed or free, and more.

```{r}
mtcars |> 
  plot2(mpg, hp, facet = starts_with("g"), facet.nrow = 2, facet.fixed_y = TRUE)
```

In this example, we limit the facet grid to two rows and ensure that all y-axes have the same scale. This simplifies `ggplot2`'s more complex `facet_wrap()` options like `scales = "free_y"` and `nrow`, which often require additional parameters.

## Leveraging Inline Transformations with `plot2()`

One of the standout features of `plot2()` is its ability to perform transformations directly within the function call. This capability can dramatically reduce the need for additional data manipulation steps, allowing you to focus on the visualisation itself. Whether you’re calculating aggregates, formatting labels, or even applying mathematical transformations, `plot2()` lets you do it all on the fly.

### Aggregations Made Easy

Suppose you want to count the number of patients admitted to each hospital. With `plot2()`, there’s no need to pre-calculate these counts; you can simply use `n_distinct()` to get the unique numbers of patients directly within the function:

```{r}
admitted_patients |> 
  plot2(x = hospital,
        y = n_distinct(patient_id))
```

This will produce a column plot showing the number of admissions per hospital, calculated directly within `plot2()`.

### Combining Data with Inline Transformations

The real power of `plot2()` comes when you start combining these inline transformations. Want to add another dimension, like the number of unique age groups within each hospital? You can do that just as easily:

```{r}
admitted_patients |> 
    plot2(x = hospital,
          y = n_distinct(patient_id),
          category = ifelse(date < "2010-01-01", "Prior to 2010", "Since 2010"),
          facet = ward,
          title = paste("Total of", n(), "patients"))
```

Here, the `ifelse()` in category adds the grouping on date, counting the number of unique patients per ward per hospital, again with no need for additional code outside of `plot2()`.

You didn’t need to `group_by()` or `summarise()` your data beforehand — `plot2()` takes care of it all, embedding the transformation directly into the plot creation process.

In `plot2()`, these tidyverse functions are available without loading other packages: `r paste0("\u0060", reexported, "()\u0060", collapse = ", ")`.

```{r}
admitted_patients |> 
  plot2(x = hospital, y = median(age), category = gender)
```

### Inline Math Transformations

You can also apply mathematical transformations directly within `plot2()`, making it easy to explore relationships in your data. For example, to plot the logarithm of patient ages across different wards, you can do this:

```{r}
admitted_patients |> 
  plot2(x = log(age), y = n(), y.transform = "log2", category = ward)
```

In a traditional `ggplot2` approach, you might have to create a new variable in your data for the logarithmic transformation (`mutate(log_age = log(age))`) before plotting, and a `scale_y_continuous(transform = "log2")` for the y axis. `plot2()` handles this seamlessly within the plotting function itself.

### Advanced Formatting on the Fly

In addition to aggregations, `plot2()` makes it easy to apply text transformations directly within the plotting function. For instance, you might want to format the date or combine different variables for the labels:

```{r}
admitted_patients |> 
  plot2(x = paste("Hospital", hospital),
        y = n(),
        category = format(date, "%Y"))
```

This example creates a plot where the x-axis labels combine the text "Hospital" with the hospital name, and the categories are based on the year extracted from the `date` variable. This level of flexibility allows for highly customised plots without the need for extra preprocessing steps.

### Combining Everything Together

Let’s put it all together in a more complex example. Suppose you want to examine the distribution of patient ages across hospitals, with each bar split by gender, and you want the x-axis to reflect the hospital name and use facets for the year of admission:

```{r}
admitted_patients |> 
  plot2(x = ifelse(gender == "F", "Females", "Males"),
        y = median(age),
        category = format(date, "%Y"),
        facet = paste("Hospital", hospital),
        x.title = "",
        y.title = "Median Patient Age",
        y.labels = function(x) paste(x, "yrs"),
        category.title = "Year")
```

In this plot, we’re using multiple inline transformations, showcasing the use of only 8 lines of code without using `dplyr` transformations manually. This level of complexity, achieved with just one `plot2()` call, demonstrates the power and flexibility of inline transformations.

## Sorting and Limiting Data

Sorting is a crucial part of data visualisation, as it helps bring clarity and focus to your plots. With `plot2()`, you have a variety of options to sort your data in different directions. Whether you want to sort by frequency, alphabetically, or by a custom order, `plot2()` provides flexible and powerful tools to get the job done.

Limiting your data helps in keeping plots clean. If you only want to display the top few items, `x.max_items` and `category.max_items`, and `facet.max_items` have got you covered.

### Basic Sorting

Let’s start with a simple sort based on frequency:

```{r}
mtcars |> 
  plot2(carb, y = n(), x.sort = "freq-desc")
```

In this example, the data is sorted by the frequency of carburettor counts in descending order. This is particularly useful when you want to highlight the most common categories in your data.

This command is akin to the functionality provided by `arrange(desc(...))` combined with `geom_bar(stat = "identity")`. However, `plot2()` condenses this into a single, intuitive step.

### Sorting Alphabetically

If you prefer to sort alphabetically, `plot2()` makes it straightforward. You can use `"asc"` or `"alpha"` to sort your data in ascending alphabetical order:

```{r}
mtcars |> 
  plot2(carb, y = n(), x.sort = "asc")
```

This command will sort the carburettor counts alphabetically, which is useful when your data naturally follows an alphabetical order.

In `ggplot2`, achieving this often involves using `factor()` levels or reordering factors manually. With `plot2()`, sorting becomes a simple matter of choosing the appropriate argument.

### Sorting with a Custom Order

Sometimes, you may want to present your data in a specific order that doesn’t follow a standard sorting method. You can define a custom order by providing a manual vector of values:

```{r}
mtcars |> 
  plot2(carb, y = n(), x.sort = c(4, 1, 6, 8))
```

In this example, the carburettor counts will be displayed in the order of 4, 1, 6, and 8, regardless of their frequency or alphabetical order. This approach is particularly useful when you want to highlight certain categories or follow a logical sequence.

In `ggplot2`, this would typically involve setting `levels` manually within a factor, which can be cumbersome. `plot2()` streamlines this by allowing custom orders directly within the `plot2()` function call.

### Limiting Data

You might not always want to show every single item in your plot. For example, if you’re dealing with a lot of categories, you can limit the display to just the most frequent ones:

```{r}
mtcars |> 
  plot2(carb, y = n(), x.max_items = 5, type = "col")
```

This limits the plot to the top 5 most frequent carburetor counts, simplifying the visualisation and making it easier to focus on the most important data.

In `ggplot2`, you might need to filter your data with `top_n()` before plotting. `plot2()` handles this within the plotting function, further reducing the need for external data manipulation.

### Combining Sorting with Limiting

Sorting can be combined with other features in `plot2()` to create more sophisticated plots. For example, you might want to sort by frequency and also limit the number of items displayed:

```{r}
mtcars |> 
  plot2(carb, y = n(), x.sort = "freq-desc", x.max_items = 5, type = "col")
```

This plot will show only the top 5 most frequent carburettor counts, sorted in descending order. This combination helps to declutter your visualisation and focus on the most significant parts of your data.

In a `ggplot2` approach, this might involve combining `arrange()`, `top_n()`, and `geom_col()`. `plot2()` consolidates these operations into a single, coherent function call.

### Visualising Sorting Options

To illustrate the impact of different sorting options, let’s create a plot that uses a custom sort order and highlights the flexibility of `plot2()`:

```{r}
admitted_patients |> 
  plot2(x = age_group, y = n(), x.sort = c("55-74", "75+", "25-54"), category = hospital, stacked = TRUE)
```

In this example, the `age_group` axis is sorted in a custom order, with the "55-74" group first, followed by "75+" and then "25-54". The bars are stacked by hospital, providing a clear comparison across the specified age groups.

This functionality parallels `ggplot2`’s ability to reorder factors using `forcats::fct_relevel()` or manually reordering levels within a factor, but with less effort and more clarity.

## Customising Colours

Customising colours in `plot2()` is intuitive and flexible. You can use pre-set colour scales like `viridis`, or define your own colours. The `colour` and `colour_fill` arguments control the appearance of your plot.

### Using Pre-set Colour Scales

For a quick and visually appealing colour scheme, you might use a `viridis` palette:

```{r}
mtcars |> 
  plot2(mpg, hp, category = cyl, colour = "viridis")
```

This applies the `viridis` colour scale, which is particularly good for making sure your plot is accessible to those with colour vision deficiencies.

This replaces the need for `ggplot2`'s `scale_colour_viridis_c()` or `scale_fill_viridis_c()`, providing a more straightforward interface.

### Defining Custom Colours

If you want more control, you can define specific colours for each category:

```{r}
mtcars |> 
  plot2(mpg, hp, category = cyl, colour = c("4" = "red", "6" = "blue", "8" = "green"))
```

You can use any colour that the implemented `get_colour()` function can understand, such as colour names and HTML codes:

```{r}
get_colour("red")
get_colour("#FF0000")
get_colour("ff0000")
get_colour("f00")
```

But most importantly, you can register new colours to the `plot2` package. Let's assume these six colours are from the style of your university/company/etc.:

```{r}
register_colour(navy_blue = "#1F3A93",
                burnt_orange = "#D35400",
                forest_green = "#2C6F47",
                goldenrod_yellow = "#DAA520",
                slate_grey = "#708090",
                plum_purple = "#8E4585")

# Then register the whole colour list too:
register_colour(OurOrganisation = c("navy_blue", "burnt_orange",
                                    "forest_green", "goldenrod_yellow",
                                    "slate_grey", "plum_purple"))
```

You can now use these colours in your plots!

```{r}
iris |>
  plot2(x = Species, y = where(is.double), colour = "OurOrganisation")
```

This approach mirrors `ggplot2`'s `scale_colour_manual()` but with an added layer of flexibility through the `register_colour()` function. This makes it easier to maintain consistent branding or theme-specific colours across multiple plots.

### Default Colours

At default, `plot2()` sets no colours, meaning that it uses ggplot2 colours. With some simple R options, it is very easy to switch to another colour set.

```{r, echo = FALSE}
plots <- function() {
  p1 <- iris |> plot2(title = "", y.title = "", x.title = "")
  p2 <- admitted_patients |> plot2(gender, n(), title = "", y.title = "", x.title = "")
  p3 <- admitted_patients |> plot2(gender, n(), ward, title = "", y.title = "", x.title = "")
  p4 <- admitted_patients |> plot2(gender, age, type = "boxplot", title = "", y.title = "", x.title = "")
  p5 <- admitted_patients |> plot2(gender, n(), age, category.character = FALSE, title = "", y.title = "", x.title = "")
  p6 <- netherlands |> plot2(datalabels = FALSE, title = "", y.title = "", x.title = "")
  p7 <- admitted_patients |> plot2(gender, age, hospital, type = "boxplot", title = "", y.title = "", x.title = "")
  p8 <- admitted_patients |> plot2(age, type = "histogram", title = "", y.title = "", x.title = "")
  (p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8)
}
```

Default:

```{r, echo = FALSE, message = FALSE, fig.height = 7.5, fig.width = 7}
options(plot2.colour = NULL, plot2.colour_sf_fill = NULL)
plots()
```

Using any [viridis colour palette](https://sjmgarnier.github.io/viridis/articles/intro-to-viridis.html):

```{r}
options(plot2.colour = "viridis")
```

```{r, echo = FALSE, message = FALSE, fig.height = 7.5, fig.width = 7}
plots()
```

```{r}
options(plot2.colour = "magma")
```

```{r, echo = FALSE, message = FALSE, fig.height = 7.5, fig.width = 7}
plots()
```

```{r}
# our previously set manual colour set:
options(plot2.colour = "OurOrganisation")
```

```{r, echo = FALSE, message = FALSE, fig.height = 7.5, fig.width = 7}
plots()
```

```{r, echo = FALSE}
# reset again
options(plot2.colour = NULL, plot2.colour_sf_fill = NULL)
```


## Advanced Plot Types: Geographies, Dumbbells and Sankey Diagrams

`plot2()` shines when it comes to more complex plot types. Let’s explore four advanced options: geography plots, UpSet plots, dumbbell plots and Sankey diagrams.

### Plotting Geometries (`sf` objects)

Objects of class `sf` (*simple feature*) are handled by `plot2()` like any other data set; `plot2()` applies spatial functions such as `geom_sf()` and `geom_sf_text()` automatically wherever needed. The built-in `theme_minimal2()` theme makes sure you get a clean map, by removing the unnecessarily verbose axes and background colour.

The included `netherlands` data set is such an `sf` object:

```{r}
plot2(netherlands)
```

`ggplot2` users would typically need to specify `geom_sf()` and `geom_sf_label()` explicitly and might spend extra time adjusting themes to remove unnecessary axes or background elements. `plot2()` simplifies this by integrating these adjustments automatically.

### UpSet Plots for Complex Set Intersections

[UpSet plots](https://en.wikipedia.org/wiki/UpSet_plot) provide a scalable and interpretable alternative to Venn diagrams for visualising complex set intersections. `plot2()` allowing the quick creation of UpSet plots to explore overlapping elements across multiple groups, revealing patterns not easily captured by traditional methods.

```{r}
df <- data.frame(var1 = sample(c(TRUE, FALSE), size = 50, replace = TRUE),
                 var2 = sample(c(TRUE, FALSE), size = 50, replace = TRUE),
                 var3 = sample(c(TRUE, FALSE), size = 50, replace = TRUE),
                 var4 = sample(c(TRUE, FALSE), size = 50, replace = TRUE),
                 some_group = sample(c("A", "B"), size = 50, replace = TRUE),
                 some_value = runif(50))
print(head(df))

df |>
  plot2(x = var1:var4,
        type = "upset",
        title = "An UpSet plot")

df |>
  plot2(x = var1:var4,
        y = some_value,
        category = some_group,
        type = "upset",
        colour = "OurOrganisation",
        y.title = "Mean Outcome",
        x.title = "Set Size",
        title = "Another UpSet plot",
        caption = "Here a caption",
        summarise_function = mean,
        x.sort = TRUE, # sort on dots: 4 first, then 3, then 2, ...
        category.sort = c("B", "A"))
```

### Dumbbell Plots for Comparisons

Dumbbell plots are excellent for comparing two categories side by side:

```{r}
admitted_patients |> 
  plot2(x = hospital,
        y = n_distinct(patient_id),
        category = gender,
        type = "dumbbell")
```

This plot type is particularly useful for showing the difference between two related metrics across different groups. In `ggplot2`, creating a dumbbell plot often involves using `geom_segment()` and `geom_point()` in combination, with precise manual adjustments for aesthetics. `plot2()` offers a straightforward, single-function alternative.

### Creating Sankey Diagrams

Sankey diagrams are another powerful visualisation, perfect for showing flow or movement from one state to another:

```{r}
Titanic |> 
  plot2(x = c(Class, Age, Survived), category = Sex, type = "sankey")
```

Sankey plots are a great way to visualise transitions, such as the survival rates of passengers on the Titanic across different classes and age groups. In `ggplot2`, Sankey diagrams require additional packages or custom code. `plot2()` makes this complex plot type accessible directly within the function.

## Adding Elements with `add_*()` Functions

Beyond the basics, `plot2()` allows you to add layers and elements to your plots with the `add_*()` functions. These are designed to integrate seamlessly with `plot2()` and `ggplot2`.

### Adding Lines, Points, and Columns

Adding a line to an existing plot is as easy as calling `add_line()`:

```{r}
p <- mtcars |> 
  plot2(mpg, hp, as.character(cyl),
        category.title = "Cylinders")

# Notice how plot2() adds 3 lines here, since `category` is set
p |> 
  add_line(y = mean(hp))

p |> 
  add_line(y = mean(hp), colour = "red", legend.value = "Average HP")
```

This adds a line at the mean horsepower. The `add_*()` functions automatically correct for `category`, removing the need for pre-processing steps to determine (in this case) the mean. A legend item can also be added by just setting `legend.value`.

In `ggplot2`, this would typically require adding `geom_hline()` with an aesthetic mapping. `plot2()` simplifies this to a single function call.

### Adding Points

Points can be added to highlight specific values or observations:

```{r}
p |> 
  add_point(x = median(mpg), y = median(hp), shape = 4, size = 10)
```

In this example, we add a point at the median `mpg` and `hp`, using a cross shape and a larger size for emphasis. This mirrors `geom_point()` in `ggplot2` but is more tightly integrated with `plot2()`’s streamlined interface.

### Adding Columns

Adding columns is just as straightforward, and can be used to create bar charts or other similar visualisations:

```{r}
p |> 
  add_col(x = cyl, y = n(), width = 0.5)
```

Here, we add columns based on the number of cylinders, making it easy to compare the counts across different groups. This replaces `geom_col()` or `geom_bar(stat = "identity")` in `ggplot2`, maintaining simplicity and directness.

### Plotting Error Bars

Plotting error bars is just as simple:

```{r}
p |> 
  add_errorbar(min = hp - 10, max = hp + 10)
```

This adds error bars to the plot, showing a range of ±10 around the `hp` values. `plot2()` abstracts away the complexities of `geom_errorbar()`, offering a more intuitive syntax.

### Adding Spatial Features with `add_sf()`

For those working with geographic data, `add_sf()` makes it easy to integrate spatial features:

```{r}
plot2(netherlands) |> 
  add_sf(netherlands, colour_fill = NA, colour = "red", linewidth = 2)
```

This example adds spatial features to a plot of the Netherlands, with an extra border around the provinces. `ggplot2` users would typically use `geom_sf()` for this purpose, but `plot2()` integrates this directly, making geographic plotting more accessible.

## Customising Fonts in `plot2()`

Another powerful feature of `plot2()` is its flexibility with fonts. Whether you’re aiming for a professional look, something playful, or anything in between, `plot2()` allows you to easily customise fonts to suit your needs.

### Using System Fonts and Google Fonts

`plot2()` supports all installed system fonts, as well as over 1,400 Google Fonts, giving you an extensive range of choices for your plots. Missing fonts will be downloaded automatically, and the required DPI will be set for you, even in R Markdown outputs.

Here’s how you can apply a custom font from Google Fonts, which will be downloaded automatically:

```{r, fig.showtext = TRUE}
mtcars |> 
  plot2(mpg, hp, wt * 1000,
        facet = ifelse(vs == 0, "V-shaped Engine", "Straight Engine"),
        font = "Rock Salt",
        title = "Custom Font Example", 
        x.title = "Miles per US gallon",
        y.title = "Gross Horsepower",
        category.title = "Weight (lbs)",
        facet.fixed_y = TRUE)
```

In this example, the `Rock Salt` font, a playful and hand-drawn style, is applied across the plot.

In `ggplot2`, using custom fonts often requires extra handling using the `showtext` package with manual adjustments in theme settings. `plot2()` simplifies this by integrating `showtext` into its core, including automatic downloading and font scaling.

### Creating Consistent Themes with Fonts

By setting font options globally through R options, you can maintain consistent styling across multiple plots. This is particularly useful when preparing a series of visualisations for a report or presentation.

```{r, fig.showtext = TRUE}
options(plot2.font = "Lobster")
options(plot2.colour = "viridis")

mtcars |> 
  plot2(mpg, hp, title = "Consistent Font Example",
        text_factor = 1.5)
```

In this example, all plots will now use the `Lobster` font, ensuring a consistent appearance across your visualisations. The `text_factor` argument shows that it can be used to scale the text size of all plot elements, ensuring readability and aesthetic balance.

```{r}
# reset again
options(plot2.font = NULL, plot2.colour = NULL)
```

Customising fonts in `plot2()` is both flexible and straightforward, allowing you to tailor the typography of your plots to match your project's style. Whether you're using a system font or one of the many Google Fonts, `plot2()` makes it easy to create visually appealing and consistent plots with minimal effort.

# Discussion

`plot2()` is more than just a wrapper for `ggplot2`; it embodies a philosophy of simplifying and streamlining the plotting process while building on the robust foundation that `ggplot2` provides. If you're familiar with `ggplot2`, you know that its explicit, detailed approach to plot creation is one of its greatest strengths, allowing for unparalleled control over every aspect of a visualisation. However, with that power comes the need for repetitive coding and a certain level of complexity that can be challenging, especially for beginners.

### Enhancing Workflow Efficiency

The primary goal of `plot2()` is to make data visualisation faster and more intuitive by automating many of the routine tasks involved in creating plots. Whether you’re working with simple data or complex datasets requiring advanced transformations, `plot2()` helps you get to the final visualisation with fewer lines of code and less cognitive load. This not only speeds up your workflow but also makes your code more readable and maintainable.

For instance, the ability to perform inline transformations directly within the plotting function eliminates the need for pre-processing steps that would otherwise require additional code blocks using, e.g., `dplyr` or `tidyr`. This feature of `plot2()` alone can drastically reduce the complexity of your code, especially when working with large and multifaceted datasets.

### A Natural Extension of `ggplot2`

`ggplot2` has established itself as a cornerstone of data visualisation in R, and for good reason. Its layer-based approach and extensive customisation options make it incredibly powerful. `plot2()` is designed to complement these strengths by offering a more streamlined interface that automates many of the common tasks in `ggplot2`. Think of `plot2()` as a natural extension of `ggplot2` — one that retains the underlying power while simplifying the process, especially for users who prefer a more direct path from data to visualisation.

The `add_*()` functions exemplify this philosophy by allowing users to add layers and elements with minimal code, maintaining the flexibility of `ggplot2` while reducing the need for repetitive boilerplate. This makes it easier to experiment with different visualisations and iterate quickly on your designs.

### Flexibility Without Complexity

One of the key advantages of `plot2()` is that it retains the flexibility of `ggplot2` while reducing the need for detailed specification. The automatic handling of plot types, axis settings, and scales means that you can create sophisticated plots with minimal input. Yet, when you need to take control, `plot2()` offers all the options you would expect, from custom sorting to precise colour control and font selection.

This balance between flexibility and simplicity is what makes `plot2()` a powerful tool for both beginners and experienced users. Beginners can quickly produce high-quality plots without getting bogged down in the details, while advanced users can still leverage the full power of `ggplot2` when necessary.

### Philosophical Underpinnings

At the heart of `plot2()` is a commitment to making data visualisation more accessible and less intimidating. The goal is to lower the barriers to entry for creating professional-quality plots, thereby empowering more people to engage with data in a meaningful way. This aligns with the broader trend in the R community and the tidyverse towards creating tools that are both powerful and easy to use.

`plot2()` is also designed with the understanding that not every user needs — or wants — to be an expert in the intricacies of `ggplot2`. By providing sensible defaults and automating common tasks, `plot2()` allows users to focus on the most important part of the visualisation process: interpreting and communicating their data.

### Reflecting on Usage Scenarios

The real value of `plot2()` becomes evident in real-world usage scenarios. Whether you’re creating a quick exploratory plot or preparing a polished visualisation for publication, `plot2()` adapts to your needs. Its versatility makes it suitable for a wide range of applications, from routine data analysis to sophisticated data storytelling.

Moreover, `plot2()` integrates seamlessly with the broader tidyverse ecosystem. It works effortlessly with other tidyverse packages, allowing you to incorporate it into your existing workflows without disruption. This ensures that you can continue to use the tools you’re familiar with while benefiting from the efficiencies that `plot2()` brings.

### Looking Forward

As the R community continues to evolve, so too will the tools we use for data visualisation. `plot2()` represents a step towards a more user-friendly and efficient future, where the focus is on insights and communication rather than technical details. It’s a tool that grows with you — whether you’re just starting out or pushing the boundaries of what’s possible with data visualisation.

In the end, the success of any tool is measured by how well it meets the needs of its users. `plot2()` has been designed with a deep understanding of the challenges and frustrations that come with creating complex plots in R. By addressing these challenges head-on, `plot2()` aims to make your data visualisation journey smoother, more enjoyable, and ultimately more productive.
